%        File: report.tex
%
\documentclass[a4paper]{article}
\usepackage[pdftex]{graphicx}
\usepackage{float}
\usepackage{hyperref}
\usepackage{amssymb}
\floatstyle{ruled}
\newfloat{listing}{hbtH}{lop}
\floatname{listing}{Listing}
\renewcommand{\topfraction}{0.85}   % This sets the percentage for how much floats get from the ‘top’ of the text of a page
\renewcommand{\textfraction}{0}   % This sets a similar percentage for how much of a page needs to be text before no more floats can be placed on that page
\renewcommand{\floatpagefraction}{0.80} % This sets how much of the page must be taken by a float before that page can be ‘all’ floats

\author{Etienne Kneuss\\
\texttt{etienne.kneuss@epfl.ch}
}
\title{Static Analysis for the PHP Language}
\bibliographystyle{unsrt}
\begin{document}
\maketitle
\begin{abstract}
  This report presents the work that was done to implement a static analysis
  tool for the PHP programming language. The kind of analysis done by the
  compiler or by the multiple development environments are very limited. This
  tool aims at providing further feedback to a developer by checking for
  multiple bug conditions or mis-uses of the language and should reduce the
  risk of encountering fatal errors at runtime after deployment of PHP based
  application.
\end{abstract}
\section{Introduction}
The PHP programming language is a compiled language. However, the compilation is
done by default each time a certain PHP file is requested. This means that the
compilation has to be fast and hence can't do much analysis and checks. Most of
the checks are thus made at runtime. The purpose of this tool is to allow the
developer to check its work for runtime-errors, without actually running its
application.
\section{Design}
This tool consists of a custom PHP compiler, translating the code
into structures such as Abstract Syntax Trees (AST) and then Control Flow Graphs
(CFG). Fixpoint analysis is then performed on the control flow graphs. This
compiler is divided into five distinct parts:
\begin{enumerate}
  \item Lexing
  \item Parsing and Abstract Syntax Trees
  \item Semantic Analysis
  \item Control Flow Graphs
  \item Type flow Analysis
\end{enumerate}

\subsection{Lexing}
The lexer uses a modified version of JFlex\footnote{JFlex: http://jflex.de}. It
is thus written in java. The original PHP compiler uses flex as scanner. The
modification done to JFlex is the implementation of flex's \verb=yymore()=,
which currently lacks in JFlex. This function is used to tell the scanner that
the following rules should be appended to \verb=yytext= instead of overwriting
it. The tokens generated are closely related to those PHP 5.3.0 would generate.
One notable difference is that we translated every single-char tokens to their
respective name.

\subsection{Parsing and Abstract Syntax Trees}
The parser, as mentioned before, is based on a modified version of CUP. CUP is
a java based parser generator which uses a syntax similar to yacc, the parser
generator used in the original PHP compiler. The version of CUP used for this
project differs from the original one in two points:
\begin{enumerate}
  \item The original CUP cannot handle large syntax files. Indeed, since CUP
    directly writes the parsing tables as properties of the generated class,
    the size of the class properties can become too large for \emph{javac}. The
    PHP grammar being quite big, the resulting parser class couldn't be compiled.
    To solve the problem, CUP was modified to provide an option specifying that
    the multiple tables should be written into files and loaded at runtime.
  \item To avoid the troubles of referencing scala classes, the parser
    generator was modified to translate the application of the production into a
    simple tree which could then be translated by a Scala class entirely.
\end{enumerate}

\subsection{Semantic Analysis}
Some checks can be done directly by looking at the AST. Indeed, PHP provides a
wide range of features, some of them being considered nowadays as bad practice.
This tool emits notices for the following issues:

\begin{itemize}
  \item Non top-level declaration: PHP allows a developer to conditionally
    declare functions or classes. This not only generates problems with the
    following analysis, but also induces some performance hits on servers equipped
    with so-called opcode cachers. Those opcode cachers are responsible for caching
    the intermediate--or compiled version of each file, function and class. The
    goal being to speed up the process by reducing the number of compilations
    required per request. This cannot be done easily if those declarations are
    conditional.
  \item Call-time pass-by-ref: a function accepting a reference\footnote{PHP
    References: https://php.net/references} has to be defined as such, but PHP
    allows the developer to pass a reference at the time of the call, even to a
    function not declared to receive one. This feature is deprecated, and can cause
    unwanted side-effects.
  \item Non-trivial include calls: the \verb=include=\footnote{PHP Include:
    http://php.net/include} statement allows a developer to execute the given file
    in the current scope. The argument representing the file to include can be
    dynamic. This tool will try to resolve dynamic expressions that are most
    commonly used to be able to extend the analysis to that file. In case the
    expression is too dynamic, that include call will be ignored. Any
    combination of those expressions are considered as trivial dynamic expressions
    for include calls:
    \begin{itemize}
      \item the concatenation operator: "."
      \item the \verb=dirname()=\footnote{PHP dirname: http://php.net/dirname} function,
          used to retrieve the parent directory of the path passed as argument
      \item constants
      \item class constants
      \item string
    \end{itemize}
    With such trivial expressions, the analyzer will look for it in the
    multiple include-paths entries and directly attach the corresponding AST to
    the current one.
  \item Dynamic object properties: PHP allows dynamic references to an object
    property using a variable or expression (e.g. 
    \verb.$name = "a"; $obj->$name. instead of \verb.$obj->a.). This is usually
    considered as bad practice since arrays are usually preferred for such
    tasks.
  \item Dynamic variables: PHP allows to reference a variable using either a
    variable, or an expression\footnote{PHP Variable variables:
    http://php.net/variables.variable}: (\verb=$$var= or
    \verb=${'prefix'.$name.foo()}=).
  \item Assignations in conditional expressions: assignations in PHP return the
    value assigned, they are hence valid expressions inside conditional
    expressions. However, history tells us that, most of the time, this is an
    actual typographic error replacing the assignation operator \verb&=& with
    the comparison operator \verb&==&. This tool will thus emit a warning if
    such expression is found inside an \verb&if()& or \verb&for()& condition.
    We exclude \verb&while()& on purpose as there is a common use-case where
    assignations are done directly inside the \verb&while()&
    expression\footnote{PHP mysql\_fetch\_assoc:
    http://php.net/mysql\_fetch\_assoc}.
\end{itemize}

Another part of the semantic analysis consists of validating the usage of
identifiers such as variables, functions or classes. The goal is to ensure that
no obvious semantic errors such as inheritance cycles, or visibility
inconsistencies exist.

\subsubsection{API Importation}
By default, PHP comes with a very dense library of functions and classes. In
fact, the main extensions that are shipped with PHP consist of more than 2'500
functions and classes. Being able to correctly represent this internal API is a
key factor to obtain useful analysis results. This API is stored in an external
XML file, allowing easy modifications and also do not require a re-compilation.
Additionally, a \verb=--apis= command line option is available to specify a
list of API files that can be imported into the symbol tables. This is
especially useful for large projects as it allows focused analysis.

\subsubsection{Annotations}
One of the language aspect that is limiting the analysis is that PHP does
not support type specifications--or type hinting for everything. In fact, it
is only accepted in function arguments, and the type specification is limited
to \emph{arrays} and \emph{objects}. As a result, the type of the return value
cannot be specified, and arguments that take scalar values such as
\emph{booleans} or \emph{integers} cannot be specified as such.

However, many project use formatted comments that they can then
render into an API documentation using some tools, such as php
documentor\footnote{phpDocumentor: http://www.phpdoc.org}:

\begin{listing}
  \begin{verbatim}
<?php
class A {
  /**
   * @var int
   */
  public $plop;

  /**
   * @param string $str
   * @return int
   */
  public function length($str) { return strlen($str); }
}

?>
\end{verbatim}
  \caption{Block comments as annotations}
\end{listing}

The analyser will look in the AST for comments directly preceding functions,
methods or properties declarations. It will then try to extract such type
informations and inject it as type hints in the AST. The analyzer will then
simply use them like normal type hints during the type checking phase.

\subsection{Control Flow Graphs}
In fact, the AST is only the intermediate step that only let us do trivial
checks. To analyse more complex aspects of a program, this tool will use
control flow graphs (CFG) which are derived from AST's.  Control flow graphs
provide a model of the flow of values through the different control structures.
Because PHP is dynamically typed, and since variables are not declared, it is
impossible to infer and check types solely based on the structural
representation of the code. For such tasks, we require CFG's to be able to
reproduce the flow of types that would happen in an execution of the code. The
main focus of this analysis will be on analysing type flows. In the CFG, each
vertex correspond to a program state between statements, and each directed edge
represents the application of a statement.

\section{Analysis}
\subsection{PHP Overview}
Before trying to see how to analyse PHP code, it is important to understand how
PHP works, this is a small overview of the features PHP provides that are
relevant to this analysis.
\subsubsection{Types}
PHP is a weakly, dynamically typed language, meaning that types are not
specially associated to variables, but to values. A variable may hence hold
values of different types in its lifetime. It is also weakly typed as it allows
automatic--or implicit type conversions when performing an operation on values
of different types.  This is specified as part of PHP as ``type
juggling''\footnote{PHP Type Juggling:
http://php.net/language.types.type-juggling}. In PHP, we have the following
types:
\begin{itemize}
  \item \textbf{Booleans}
  \item \textbf{Integers}
  \item \textbf{Floating point numbers}
  \item \textbf{Strings}
  \item \textbf{Arrays}: Arrays in PHP are ordered hashmaps, allowing either
    strings or integers as keys.  They can contain values of any type that may
    be mixed.
  \item \textbf{Objects}
  \item \textbf{Resources}: Resources represent special data types such as file
    handle, database connection links. They can be of different types. PHP
    extensions that define resource types are responsible of handling them
    appropriately.
  \item \textbf{Null}: This is the default type to any undefined or
    uninitialized variable.
\end{itemize}

\subsubsection{Object Oriented Programing}
PHP supports OOP as of PHP4, but many features have later been added into PHP5.
The OO model as well as the syntax used is closely related to a subset of what
Java offers: public/protected/private visibility, single inheritance, with the
support of interfaces. PHP provides object properties, object methods, static
properties, static methods and class constants. You can dynamically define an
object property in PHP, its visibility will default to public. However, you
cannot dynamically define methods, static members, or constants.

\subsubsection{Limitations on Analysis}
Many of the dynamic features available in PHP will get in the way of a sound
analysis.  Here is a non-exhaustive list of such features:
\begin{itemize}
  \item \textbf{autoload}\footnote{PHP Autoload: http://php.net/autoload}:
    PHP allows you to trigger a function call in case an undefined class was
    used.  This function call is usually used to subsequently load the
    appropriate class at runtime. This feature allows programmers to only load
    classes on demand.
  \item \textbf{\_\_call}\footnote{PHP Overloading: http://php.net/\_\_call}:
    If a method call is defined in an object, calling an undefined method of
    that object will instead call the \verb&__call& method, with the name of
    the original method, and the arguments used.
  \item \textbf{\_\_callStatic}:
    This is similar to \verb&__call& but works for undefined static method
    calls.
  \item \textbf{\_\_get}\footnote{PHP Overloading: http://php.net/\_\_set}:
    This method will get called in case an access to an undefined object
    property is done. The value returned by the method will correspond to the
    property value.
  \item \textbf{ArrayAccess}\footnote{ArrayAccess Interface: http://php.net/arrayaccess}:
    The ArrayAccess interface allows an object to be used as an array. For
    example, if we have \verb&class Foo implements ArrayAccess&, then
    \verb&$foo = new Foo; echo $foo['index']& is accepted.
  \item \textbf{Dynamic accesses}
    PHP allows multiple ways to dynamically access variables, classes, members\ldots
    Here are a couple of examples:
    \begin{itemize}
      \item \verb/$$n/:
        Accesses the variable named by the value of \verb/$n/.
      \item \verb/$$$n/:
        Accesses the variable named by the value of \verb/$$n/.
      \item \verb/new $n()/:
        Constructs an object of the class named by the value of \verb/$n/.
      \item \verb/$c::$n()/:
        Call the static function named by the value of \verb/$n/ on the class
        named by the value of \verb/$n/.
      \item \verb/$n()/:
        Call the function named by the value of \verb/$n/.
    \end{itemize}
  \item \textbf{eval}\footnote{PHP Eval: http://php.net/eval}:
    The \verb&eval()& construct allows a developer to evaluate the code passed
    in as a string.
\end{itemize}

\subsection{Typeflow Analysis}
As said earlier, the main focus of this analysis tool is on types. Even though
PHP is dynamically typed, and provides very few indications of types. It is
still interesting to try to infer types and check for type safety.

\subsubsection{Type model}
We now provide a detailed description of every elements of our type model:
\begin{itemize}
  \item \emph{TAny} ($\top$): Top of our lattice, represents any type
  \item \emph{TNone} ($\bot$): Bottom of our lattice, represents no type
  \item \emph{TBoolean}: Any boolean type.
  \item \emph{TFalse}, \emph{TTrue} : True/False types, used for conditional
    type filtering (See \ref{cond-type-filtr}).
  \item \emph{TInt}: Integers
  \item \emph{TFloat}: Floating point numbers
  \item \emph{TResource}: Resources. We make no distinction between resources
    of different types. This could be improved in the future.
  \item \emph{TNull}: Null value.
  \item \emph{TString}: Strings.
  \item \emph{TUnion}: Union of multiple distinct types, represented as the
    infix $\cup$ operator. The result of this union of types depends on the operands,
    it will be described into further details in its own section.
  \item Arrays: Arrays are represented using a quite precise model, it is
    structured as follows:
    \begin{itemize}
      \item A set of $(\emph{index} \rightarrow \emph{type})$ associations for
          every well-defined array entries. The rationale behind having such a precise
          representation is that PHP arrays are often used as containers for
          multiple named values. Unlike other languages like C, PHP allows values
          in arrays to be of multiple types. Those values are then used
          independently. It is thus important to keep track of the types
          associated to each index.
        \item An optional array-wise default type. This type represents the
          possible contamination of an array via a dynamic access. This type is
          called the ``polluted type''. For example, the code
          \verb/$a[$foo] = 2;/ will have the following effect on a well-defined
          array: $\forall (e \rightarrow t) \in entries: t = t \sqcup
          \emph{TInt}$. Also, we have $\emph{polluted type} = \emph{polluted
          type} \sqcup \emph{TInt}$. We can also represent this default type as
          $(? \rightarrow \emph{type})$. A side effect of having a non-empty
          polluted type will be to turn off any notices related to undefined
          entries. In fact, any lookup for an undefined entry in an array will
          default to the polluted type, if any, or issue a notice for a
          potentially undefined index.
    \end{itemize}
    We note that, even though PHP are \emph{ordered} hashmaps, we totally
    ignore to model the order as it is not something that people commonly rely
    upon. Additionally, we argue that if somebody is traversing an array in
    order, potentially all types of the array will get visited. Hence,
    iterating over an array will yield values of type equal to the union of all
    possible types of the array.

    Two types are used internally to represent arrays:
    \begin{itemize}
      \item \emph{TArray}: The general type representing the model described
        above.
      \item \emph{TAnyArray}: This represents an array being the
        supertype of all arrays. It is in fact an alias to an array whose
        only entry is $(? \rightarrow \emph{TAny})$.
    \end{itemize}
  \item Objects: Objects are represented differently than other types. The
    reason is that when passing, assigning or returning objects PHP will pass
    the object reference. For example, in 
    \verb/$a = new Foo; $b = $a; $a->foo = 2/, both \emph{\$a} and \emph{\$b}
    represent the same object, hence both variables are affected by the object
    modification. We need to reproduce that level of indirection as part of our
    type model for objects. This is done by introducing object references as
    types, mapping the real object in an object store. That way the type can
    be copied around and propagated to different values and still point to the
    same real object type. We thus have two different levels of types:
    \begin{itemize}
          \item \textbf{Object references:} This is the type to which
            any object value is assigned.
          \item \textbf{Real objects:} The real object is pointed to by the
            object reference. This is the actual object type storing mappings to
            methods, properties and constants. It is stored in the object store
            and is not directly exposed as types of values.
    \end{itemize}

    Real objects are decomposed into two object types:
    \begin{itemize}
          \item \emph{TRealClassObject}: Representing an object of a
            defined class. This type will have, like arrays, a index-type
            mapping for properties. Since object properties can be dynamically
            defined, we also have a so-called polluted type for properties.
          \item \emph{TRealObject}: This is very similar to
            \emph{TRealClassObject}, the only difference is that this object
            carries no class information. It is mostly used for type
            constructions. For instance, in the code
            \verb/$a = new A;/ \verb/$a->b = 2;/ the statement 
            \verb/$a->b = 2;/ will merge the type of \verb/$a/ which is
            $\emph{TRealClassObject}(A)(\cdots)$ with a constructed
            $\emph{TRealObject}(a \rightarrow \emph{TInt})$. This will result
            in  $\emph{TRealClassObject}(A)(\cdots, a \rightarrow
            \emph{TInt})$.
    \end{itemize}
\end{itemize}

\subsection{Type Lattice}
A lattice is usually defined by three things: a partial order relation
($\sqsubseteq$), a join operation ($\sqcup$), and a meet operation ($\sqcap$). As noted
earlier, the $\sqsubseteq$ relation is the subtyping relation, we have that $A \sqsubseteq B
\Leftrightarrow A <: B$. The join operation $A \sqcup B$ returns the smallest
type $C$ such that $A \sqsubseteq C \wedge B \sqsubseteq C$. As for the meet operation
we can ignore it in our case since it is not necessary for top-down analysis.

\subsubsection{Definition of $\sqsubseteq$}
Those are, in order, the rules that apply to determine whether $t_1 \sqsubseteq t_2$:
\begin{enumerate}
  \item $\forall t_2 \in types : \emph{TNone} \sqsubseteq t_2$
  \item $\forall t_1 \in types : t_1 \sqsubseteq \emph{TAny}$
  \item $\emph{TFalse} \sqsubseteq \emph{TBoolean}$
  \item $\emph{TTrue} \sqsubseteq \emph{TBoolean}$
  \item $\forall (t_1:  TObjectRef): t_1 \sqsubseteq \emph{TAnyObject}$
  \item $\forall (t_1,t_2: TObjectRef): t_1 \sqsubseteq t_2$
    if all the following rules apply ($o_1 = store(t_1.ref), o_2 = store(t_2.ref)$):
    \begin{enumerate}
        \item $o_1.class <: o_2.class$ if $o_1$ and $o_2$ are real class objects
        \item $o_1.pollutedType \sqsubseteq o_2.pollutedType$ if any
        \item $\forall f \in o_3.fields: f \in o_1.fields \wedge o_1.fields(f)
          \sqsubseteq o_2.fields(f)$
    \end{enumerate}
  \item $\forall (a: \emph{TArray}) : a \sqsubseteq \emph{TAnyArray}$
  \item $\forall (a_1,a_2: \emph{TArray}) : a_1 \sqsubseteq a_2$
    if all the following rules apply:
    \begin{enumerate}
        \item $a_1.pollutedType \sqsubseteq a_2.pollutedType$ if any
        \item $\forall e \in a_2.entries: e \in o_1.entries \wedge a_1.entries(e)
          \sqsubseteq a_2.entries(e)$
    \end{enumerate}
  \item $\forall (u_1,u_2: \emph{TUnion}) : u_1 \sqsubseteq u_2 \Leftrightarrow \forall t_2 \in u_2: \exists t_1 \in u_1: t_1 \sqsubseteq t_2$
  \item $\forall t_1 \in types, (u_2: \emph{TUnion}) : t_1 \sqsubseteq u_2 \Leftrightarrow \forall t_2 \in u_2: t_1 \sqsubseteq t_2$
  \item $\forall (u_1: \emph{TUnion}), t_2 \in types : u_1 \sqsubseteq t_2 \Leftrightarrow \forall t_1 \in u_1: t_1 \sqsubseteq t_2$
  \item $\forall t_1,t_2 \in types : t_1 \nleq t_2$ If none of the rules above applies
\end{enumerate}

\subsubsection{Definition of Join ($\sqcup$)}
Those are, in order, the rules that apply to determine the result of $t_1 \sqcup t_2$:
\begin{enumerate}
  \item $\forall t \in types : t \sqcup TNone = \emph{TNone} \sqcup t = t$
  \item $\forall t \in types : t \sqcup t = t$
  \item $\forall (t: \emph{TObjectRef}): t \sqcup \emph{TAnyObject} = \emph{TAnyObject} \sqcup t = \emph{TAnyObject}$
  \item $\forall (t_1,t_2: \emph{TObjectRef}): t_1 \sqcup t_2 = TUnion(t_1 t_2)$. As we
    know that at this point, the types must be different, which means they
    point to different objects.
  \item $\forall (t: \emph{TArray}): t \sqcup \emph{TAnyArray} = \emph{TAnyArray} \sqcup t = \emph{TAnyArray}$
  \item $\forall (t_1,t_2: \emph{TArray}): t_1 \sqcup t_2 = $ an array where the
    polluted type is the join of both polluted types, and the entries is the
    union of both entry sets. Note that for an entry present in both arrays,
    the join of its types is taken.

  \item $\forall t_1,t_2 \in types: t_1 \sqcup t_2 = \emph{TUnion}(t_1, t_2)$. If none of the above applies, we get a union of types.
\end{enumerate}

\subsection{Analysis Mechanics}
As we know, this analysis is based on point-wise lattices. A lattice is a
structure providing a partial order between its elements. This is specially
useful to represent types (the order relation being subtyping) which is what
we use in our case.

The role of a point-wise lattice is to keep track of a lattice value, or type in
our case, for every values. We will denote those point-wise lattice elements as
type environments. A type environment is basically a mapping between values and
types.

The analysis algorithm starts by assigning a base type environment to every CFG
vertexes, or a global bottom ($\bot_g$) value, indicating whether such a vertex
has been visited yet. Then, the algorithm starts with the entry of the CFG, and
will visit the graph in order. For each edge $e = v_1 \rightarrow^s v_2$ where
$s$ is the statement, we can calculate the environment at $v2$ coming from
$e_i$ by applying a transfer function on $v_1$ with $s$. This will represent
the type environment in which $v_2$ will be in case that edge $e$ was taken to
get to $v_2$. Of course, one vertex may have multiple entry edges and so, to
calculate the actual type environment of $v_2$ we will have to join every
environment $env_i$ defined as $env_i = transferFunction(v_i, s_i)$ where $(v_i
\rightarrow^{s_i} v_2) \in edges$. This type environment join operator
($\cup_e$) is defined as follows:
\begin{itemize}
  \item $\bot_g \cup_e E_i = E_i$ and $E_i \cup_e \bot_g = E_i$
  \item For $E_i \cup_e E_j$ where $E_i \neq \bot_g$ and $E_j \neq \bot_g$, we
    construct $E_u = E_i \cup_e E_j$ using the following set of rules: First,
    we define $E_i(v_j)$ as the type associated to the value $v_j$ in the
    environment $E_i$.  We have that $ \forall v_k \in values(E_i) \cup
    values(E_j):$
    \begin{eqnarray*}
      E_u(v_k) = \left \{ \begin{array}{l l}
        E_i(v_k) \sqcup E_j(v_k) & v_k \in values(E_i)    \wedge v_k \in values(E_j) \\
        E_i(v_k) \sqcup TNull    & v_k \in values(E_i)    \wedge v_k \notin values(E_j) \\
        TNull    \sqcup E_j(v_k) & v_k \notin values(E_i) \wedge v_k \in values(E_j)
      \end{array} \right \}
    \end{eqnarray*}
    Note that the join operator $\sqcup$ used with two types represents the
    \emph{join} operation on our type lattice, as described earlier in the 
    type lattice section.
\end{itemize}

We see that traversing the CFG only once will not be sufficient. For instance, in
an loop, in the first vertex in the loop, the entry edge coming from the end
of the loop is not known when we enter the loop the first time. Multiple passes
are hence required! In order for this algorithm to terminate, we check whether
we actually changed the environment. This so called ``fix-point'' algorithm will
stop whenever no change has been made to any environment during a complete
traversal of the graph. By design, the type lattice join operation will not
generate types that will always change, as it can only go ``up'' in our
lattice. However, we need to take special care in the type transfer function that
it doesn't generate differences that would jeopardize the termination. For instance
\verb/$a = new MyClass/ cannot be made to create a new object reference for
every traversal! To solve this case, we set the reference value to the program point.

\subsubsection{Conditional Type Filtering}
\label{cond-type-filtr}
So far, the analysis features are quite common and could be applied to most
languages. This is however insufficient to give a good analysis. The reason is
that most internally defined PHP functions will return \verb/false/ or
\verb/null/ in case of error. For a function normally returning an integer,
this means that any return value of this function should be typed
$\emph{TInt} \cup \emph{TBoolean}$. However, this would lead to massive amount of false
positives. Consider the following code example which can be seen in most
procedural scripts dealing with mysql queries:

\begin{listing}
  \begin{verbatim}
<?php
$result = mysql_query(..);
if ($result) {
    while($row = mysql_fetch_assoc($result)) {
      ..
    }
}
?>
  \end{verbatim}
  \caption{Fetching query results}
\end{listing}

\verb/mysql_query/ either return a result resource or false in case the query
failed. \verb/mysql_fetch_assoc/ takes a resource, and return an array
representing the row, unless it reached the end of the result set in which case
it will return false. Naively checking types would give us many false
positives:
\begin{itemize}
  \item \verb/mysql_fetch_assoc/'s argument would be of type
    $\emph{TBoolean} \cup \emph{TResource}$, which would fail with a type
    mismatch saying that it expects a \emph{TResource}.
  \item Inside the while body, \verb/$row/ would still be of type $\emph{TBoolean}
    \cup \emph{TArray}$, generating a false positive for any usage of
    \verb/$row/ as an array.
\end{itemize}

The solution to this problem is in two parts. First of all, it was necessary to
improve the precision of the boolean type and introduce \emph{TTrue} and
\emph{TFalse}.  That was, a function can be described as returning
$\emph{TResource} \cup \emph{TFalse}$.  From those types, it becomes quite
trivial to filter incompatible values when taking a branch.

For instance, with \verb/if ($result)/ the CFG will create two branches, one
for $\$result = true$ and one for $\$result \neq true$. In the first branch, we
can filter out types of \verb/$result/ that represent strictly false values,
which includes \emph{TFalse} and \emph{TNull}. The same applies for the second
branch, in which we filter out strictly true values, namely \emph{TTrue} and
\emph{TResource}. In our previous example, the type of \verb/$result/ inside
the if body would thus be filtered to \emph{TResource} which is correct.
Moreover, it will also filter assigns as conditionals, meaning that in 
\verb/while($row = mysql_fetch_assoc($result))/, the while body will have
\emph{TResource} for type of \verb/$row/. In summary, this simple filtering
trick allows for precise prototyping with reduced false-positives!

\section{Limitations and Future Work}
Most serious PHP applications out there are often based on some heavy frameworks
that may require a lot of ``magic`` features to work. Analysing such projects
would currently yield a flood of false-positive, totally hiding any potential
real problem.

Lots of features are not completely checked or integrated in the analysis,
such as exceptions, namespaces, or closures. In general, there are still
lot of areas of the language for which the analysis is not done or hasn't
been properly tested.

This tool still fails at analysing one ``core'' feature of PHP, which is
\emph{references}. As of PHP5, the usage of references is in most case obsolete
or even harmful. However, it is still used by lots of projects, for reasons
such backward compatibility or performance improvements.

\section{Conclusion}
Doing precise analysis on PHP is hard, and that is because the language has never
been developed with such concerns in mind. We can however see the potential
applications of such tool, like direct integration as part of an IDE or
plugged to a source repository as a quality assurance tool. The fact that only
few developments are aimed at providing analysis for PHP can possibly be
explained by its difficulty. As a consequence, the PHP world still has a lot
of room for analysis tools such as the one described here, which  makes its
development quite interesting and motivating.


\addcontentsline{toc}{section}{References}
\begin{thebibliography}{99}
  \bibitem{tigerbook}A.W. Appel {\it Modern Compiler Implementation in Java}
    Cambridge University Press (2002)
  \bibitem{tapl}Benjamin C. Pierce {\it Types and Programming Languages} The
    MIT Press (2002)
  \bibitem{aliasanal}N. Jovanovic, C. Kruegel, E. Kirda {\it Precise Alias
    Analysis for Static Detection of Web Application Vulnerabilities} Technical
    University of Vienna
  \bibitem{diminfer}S. Hangal, M.S. Lam {\it Automatic Dimension Inference and
    Checking for Object-Oriented Programs} Standford University
  \bibitem{recdescent}\mbox{http://lara.epfl.ch/dokuwiki/compilation:}
    \mbox{example\_efficient\_code\_for\_conditionals}
\end{thebibliography}
\end{document}
